---
title: "Chess Tournament Results"
author: "Johnny Maldonado"
date: "9/22/2026"
output: html_document
---

## Introduction

The purpose of this project is to transform a semi-structured text file containing chess tournament results into a clean dataset that can be exported as a CSV file. The final dataset will include each player's name, state, total points, pre-tournament rating, and the average pre-tournament rating of their opponents.

To complete the project, I will extract the relevant player information from the text file and organize it into a structured format. The opponent numbers listed in each round will be matched back to the corresponding players so their pre-tournament ratings can be identified. I will then calculate the average opponent pre-tournament rating for each player, verify the results, and export the completed dataset as a CSV file.

### Libraries
Libraries TBD
```{r library imports}

```

### Data import
import .txt file here 
```{r data import}

```


