---
title: "Chess Tournament Result"
author: "Johnny Maldonado"
date: "9/22/2026"
format: html
editor: visual
---
## Introduction

The purpose of this project is to transform a semi-structured text file containing chess tournament results into a clean dataset that can be exported as a CSV file. The final dataset will include each player's name, state, total points, pre-tournament rating, and the average pre-tournament rating of their opponents.

To complete the project, I will extract the relevant player information from the text file and organize it into a structured format. The opponent numbers listed in each round will be matched back to the corresponding players so their pre-tournament ratings can be identified. I will then calculate the average opponent pre-tournament rating for each player, verify the results, and export the completed dataset as a CSV file.

### Libraries

The following libraries will be used to clean, transform, and organize the tournament data.

```{r library import}
library(dplyr)
library(stringr)
library(tidyr)
library(readr)
```

```{r data import}
tournament_url <- "https://raw.githubusercontent.com/jmald1987/DATA607_Chess_Tournament_Results/refs/heads/main/tournamentinfo.txt"

tournament_data <- read_lines(tournament_url)

head(tournament_data)
```
### Extract Player Records

The tournament file contains multiple types of rows, including headers, separator lines, player information, and rating information. The player rows can be identified because they begin with a player number, while the following information row begins with the player's state.

```{r extract-players}
player_lines <- tournament_data[str_detect(tournament_data, "^\\s*\\d+\\s*\\|")]
rating_lines <- tournament_data[str_detect(tournament_data, "^\\s*[A-Z]{2}\\s*\\|")]

length(player_lines)
length(rating_lines)
```
The rows can be separated using the pipe character as the delimiter.
```{r player-rows}
player_parts <- str_split_fixed(player_lines, "\\|", 11)

players <- tibble(
  player_num = as.integer(str_trim(player_parts[, 1])),
  player_name = str_trim(player_parts[, 2]),
  total_points = as.numeric(str_trim(player_parts[, 3])),
  round_1 = str_trim(player_parts[, 4]),
  round_2 = str_trim(player_parts[, 5]),
  round_3 = str_trim(player_parts[, 6]),
  round_4 = str_trim(player_parts[, 7]),
  round_5 = str_trim(player_parts[, 8]),
  round_6 = str_trim(player_parts[, 9]),
  round_7 = str_trim(player_parts[, 10])
)

head(players)
```


The second row for each player contains the player's state and rating information. The pre-tournament rating appears immediately after

```{r parse state and pre-ttournament rating}
rating_parts <- str_split_fixed(rating_lines, "\\|", 11)

players <- players %>%
  mutate(
    state = str_trim(rating_parts[, 1]),
    pre_rating = as.integer(
      str_match(rating_parts[, 2], "R:\\s*(\\d+)")[, 2]
    )
  )

players %>%
  select(player_num, player_name, state, total_points, pre_rating) %>%
  head()
```
Each round result contains the result of the game followed by the opponent's player number. The opponent numbers will be extracted and converted into a long format so that they can be matched to the opponent's pre-tournament rating.
```{r opponents}
opponents <- players %>%
  select(player_num, starts_with("round_")) %>%
  pivot_longer(
    cols = starts_with("round_"),
    names_to = "round",
    values_to = "result"
  ) %>%
  mutate(
    opponent_num = as.integer(str_extract(result, "\\d+"))
  ) %>%
  filter(!is.na(opponent_num))

head(opponents)

```


The opponent number can now be matched to the corresponding player's pre-tournament rating.

```{r opponent ratings}
opponents <- opponents %>%
  left_join(
    players %>%
      select(
        opponent_num = player_num,
        opponent_pre_rating = pre_rating
      ),
    by = "opponent_num"
  )

head(opponents)
```
### Calculate Average Opponent Rating

The average pre-tournament rating of each player's opponents is calculated using the opponents they actually played.

```{r average opponent rating}
average_opponent_rating <- opponents %>%
  group_by(player_num) %>%
  summarise(
    avg_opponent_pre_rating = round(
      mean(opponent_pre_rating, na.rm = TRUE)
    )
  )

head(average_opponent_rating)
```
### Final Dataset

The calculated opponent averages are joined back to the player information. Only the fields required for the final CSV file are retained.

```{r chess tournament results}
final_results <- players %>%
  left_join(
    average_opponent_rating,
    by = "player_num"
  ) %>%
  transmute(
    Player_Name = player_name,
    State = state,
    Total_Points = total_points,
    Pre_Rating = pre_rating,
    Average_Opponent_Pre_Rating = avg_opponent_pre_rating
  )

final_results

```

## Verification

The assignment provides Gary Hua as an example. His pre-tournament rating is 1794 and his opponents' average pre-tournament rating should be approximately 1605.


```{r verification}
final_results %>%
  filter(Player_Name == "GARY HUA")
```

```{r export csv}
write_csv(
  final_results,
  "chess_tournament_results.csv")
```

## Conclusion
This project transformed a semi-structured chess tournament text file into a clean and structured dataset. Player information and opponent numbers were extracted from the original file, and the opponent numbers were used to match each opponent with their pre-tournament rating. These ratings were then used to calculate the average opponent rating for each player. The final dataset contains the required player name, state, total points, pre-tournament rating, and average opponent pre-tournament rating and was exported as a CSV file.